---
title: "Rcpp training"
output:
  bookdown::html_document2:
    #base_format: rmarkdown::html_vignette
    #highlight: tango
    number_sections: true
    toc: true
    #toc_float: true
    fig_caption: yes
link-citations: yes
pkgdown:
  as_is: true
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteIndexEntry{Rcpp training}
 %\VignetteEncoding{UTF-8}
urlcolor: Maroon
---

```{r options, include=FALSE}
library(knitr)
library(kableExtra)
library(pkgdown)
library(dplyr)
opts_chunk$set(echo=TRUE, cache=TRUE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```

# RcppGSL example

## C++ code
```{Rcpp RcppGSL-my_rnorm}
#include <Rcpp.h>
#include <gsl/gsl_rng.h>
#include <gsl/gsl_randist.h>
using namespace Rcpp;
// [[Rcpp::depends(RcppGSL)]]
// [[Rcpp::export]]
Rcpp::NumericVector my_rnorm(int nsamp, double mu,
                             double sigma) {
	gsl_rng *s = gsl_rng_alloc(gsl_rng_mt19937); // Random seed
	Rcpp::NumericVector beta(nsamp);
	for (int i = 0; i < nsamp; i++) {
		beta[i] = mu + gsl_ran_gaussian(s, sigma); // Random draw
	}
	return beta;
}
```
\vspace{0.3cm}

## R code
```{r r-my_rnorm, fig.show="asis", out.width="70%"}
library(Rcpp)
library(RcppGSL)
beta <- my_rnorm(100, 5, 2)
par(cex=2)
hist(beta)
```

# RcppArmadillo example

## C++ code
```{Rcpp RcppArma-example}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::export]]
arma::mat arma_scale(const arma::mat& X) {
    int n = X.n_rows, k = X.n_cols;
    arma::rowvec col_means = arma::mean(X,0); // means 
    arma::rowvec col_sd  = arma::stddev(X,0); // standard deviations
    arma::mat X_scaled(n,k);
    for (int  p= 0; p < k; p++) {
    X_scaled.col(p) = (X.col(p)-col_means(p))/col_sd(p);
    }
    return X_scaled;
}
```

## R code
```{r r-RcppArma-example}
library(Rcpp)
library(RcppArmadillo)

# Center and reduce a matrix 
X <- matrix(rnorm(50),ncol=5)
X_scaled <- arma_scale(X)
colMeans(X_scaled)
var(X_scaled)
```


# Distance computation 

## Mathematical definition

The Euclidean distance between two points whose coordinates are $A=(x_A, y_A)$ and $B=(x_B,y_B)$ is given by $$\sqrt{(x_B-x_A)^2 + (y_B-y_A)^2}.$$

## C++ function 
(in file `Code/arma_distmat.cpp`)
```{Rcpp arma_distmat}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
using namespace Rcpp;
using namespace arma;

// [[Rcpp::export]]
arma::mat arma_distmat(const arma::mat& x) { 
  int np = x.n_rows;
  arma::mat distmat; distmat.zeros(np, np);
  for (int i = 0; i < np; i++) {
    arma::vec p0 = x.row(i).t();
    for (int j = i + 1; j < np; j++) {
      arma::vec p1 = x.row(j).t();
      arma::vec diff = p0 - p1;
      double squared_diff = as_scalar(diff.t() * diff);
      distmat(j, i) = distmat(i, j) = sqrt(squared_diff);
    }
  }
  return distmat;
}
```

## R function
```{r R_distmat}
R_distmat <- function(X){
  np <- nrow(X)
  distmat <- matrix(0,nrow=np,ncol=np)
  for (i in 1:(np-1)) {
    p0 <- X[i,]
    for (j in (i+1):np){
    p1 <- X[j,]
    diff <- p0-p1
    squared_diff <- t(diff)%*% diff
    distmat[j,i] <- distmat[i,j] <- sqrt(squared_diff)
    }
  }
 return(distmat)
}
```

## Comparison of compilation times
```{r dist-benchmark}
library(rdist)
# Data simulation 
coords <- matrix(runif(2000,0,100),ncol=2)
# Call of C++ function
Rcpp::sourceCpp("Code/arma_distmat.cpp")
# Benchmark
library(rbenchmark)
Benchmark <- benchmark(
  "arma_distmat" = {arma_distmat(coords)},
  "cdist" = {cdist(coords,coords)},
  "R_distmat" = {R_distmat(coords)},
  replications=30,
  columns = c("test", "elapsed", "relative"))
```

```{r dist-benchmark-results, echo=F}
library(dplyr)
knitr::kable(Benchmark,digits=1, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"),
		                          full_width=FALSE,position = "left")
```

## Representation of results 
\vspace{0.5cm}

**1000 points dispersed all over the space : **
\vspace{0.2cm}
```{r points-representation, echo=F,  fig.show="asis", out.width="60%",out.height="60%", fig.align="left"}
# Spatial representation of points  
coords <- matrix(runif(2000,0,100),ncol=2)
plot(coords[,1],coords[,2],pch=4,xlab="x", ylab="y", main="Spatial repartition of points", col="forestgreen")
```
\vspace{0.2cm}
**Coordinates of 5 points**
\vspace{0.1cm}
```{r coordinates, echo=F}
knitr::kable(coords[1:5,],col.names = c(" x "," y "), digits=1, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"),
		                          full_width=FALSE,position = "left")
```
\vspace{0.2cm}
**Distance matrix for this points : **
\vspace{0.1cm}
```{r distance-matrix, echo=F}
distmat <- arma_distmat(coords)
knitr::kable(distmat[1:5,1:5], digits=1, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"),
		                          full_width=FALSE,position = "left")
```

# Simple linear regression 

## Mathematical definition 
\vspace{0.2cm} 
We have observations of a response variable available for $n$ individuals $Y=(y_i)_{i=1,\ldots,n}$   

and $p$ explanatory variables $(X_1,\ldots,X_p)$ such as   $X_1=(X_{11},\ldots,X_{i1},\ldots,X_{n1})'$.  

\vspace{1cm}
We want to estimate $\beta=(\beta_0,\beta_1,\ldots,\beta_p)$ such as : 
$$ y_i = \beta_0 + \beta_1X_{i1}+\ldots+\beta_pX_{ip}+\epsilon_i,$$
where $\epsilon_i \sim \mathcal{N}(0,\sigma^2)$ for $i=1,\ldots,n$.  
\vspace{0.4cm}

This gives in matrix writing :  

$$Y = X\beta + \epsilon $$ 
where $X=(\mathbb{1}_n,X_1,\ldots,X_p)$ and $\epsilon=(\epsilon_i)_{i=1,\ldots,n}$.  
\vspace{1cm}

According to the Ordinary Least Squares (OLS) method, $\beta$ is estimated by :
$$\widehat{\beta}=\left(X'X\right)^{-1}X'y.$$
\vspace{0.2cm}
We define the residuals  $$\widehat{\epsilon} = Y - X\widehat{\beta}=Y-\widehat{Y}.$$
\vspace{0.2cm}
Then residual variance is given by $$\widehat{\sigma^2}=\dfrac{\sum\limits_{i=1}^n \widehat{\epsilon_i}^2}{n-p}.$$  
\vspace{0.2cm}
Finally the variance–covariance matrix of coefficients is estimated by $$V_{\widehat{\beta}}=\widehat{\sigma^2}\left(X'X\right)^{-1}.$$
\vspace{1cm}

## C++ function
```{Rcpp fastLm}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::export]]
Rcpp::List arma_fastLm(const arma::mat& X, const arma::colvec& y) {
    int n = X.n_rows, k = X.n_cols;
        
    arma::colvec coef = arma::inv(X.t()*X)*X.t()*y;    // fit model y ~ X
    arma::colvec res  = y - X*coef;           // residuals
    // std.errors of coefficients
    double s2 = arma::sum(res.t()*res);
    arma::colvec std_err = arma::sqrt(s2*arma::diagvec(arma::inv(X.t()*X)));
    return Rcpp::List::create(Rcpp::Named("coefficients") = coef,
                        Rcpp::Named("stderr") = std_err,
                        Rcpp::Named("residuals")  = res,
                        Rcpp::Named("df.residual")  = n - k);
}
```

## R function
```{r r-fastLm}
R_fastLm <- function(X, y) {
n <- nrow(X)
k <- ncol(X)

# fit model y ~ X       
coef <-solve(t(X) %*% X) %*% t(X) %*% y
# residuals
res <- y - X %*% coef
# std.errors of coefficients
s2 = sum(t(res) %*% res)/(n - k);
std_err = sqrt(s2 %*% diag(solve(t(X)%*%X)))
return(list("coefficients" = coef,
            "stderr" = std_err,
            "residuals" = res,
            "df.residual" = n - k))
}
```

## Comparison of compilation times
```{r lin-benchmark}
library(datasets)
# Trees data-set
y <- log(trees$Volume)
X <- cbind(1, log(trees$Girth), log(trees$Height))
# Call C++ function
Rcpp::sourceCpp("Code/arma_fastLm.cpp")
# Benchmark
library(rbenchmark)
Benchmark <- benchmark(
  "arma_fastLm" = {arma_fastLm(X,y)},
  "R_fastLm" = {R_fastLm(X,y)},
  replications=100,
  columns = c("test", "elapsed", "relative"))
```

```{r lin-benchmark-results, echo=F}
knitr::kable(Benchmark,digits=4, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"),
		                          full_width=FALSE,position = "left")
```

## Representation of results 
```{r lin-results,fig.show="asis", out.width="50%",out.height="50%"}
# Trees data-set
y <- log(trees$Volume)
X <- cbind(1, log(trees$Girth), log(trees$Height))
# fit model y ~ X
arma_mod <- arma_fastLm(X,y)
R_mod <- R_fastLm(X,y)
# Comparison of residuals obtained with R and C++ functions
plot(arma_mod$residuals,R_mod$residuals,xlab="residuals with C++",ylab="residuals with R",pch=4, main="Residuals")
abline(a=0,b=1, col="red")
# Representation of the linear regression
plot(X%*%arma_mod$coefficients,y,xlab="Xbeta",ylab="y",pch=4, main ="Linear regression")
abline(a=0,b=1, col="red")
```

# Loglikelihood computation 

## Mathematical definition

Implement a function in C++ using RcppArmadillo to compute the log-likelihood corresponding to each of the following models fitted on the trees dataset used previously :  

- Model 1 : $y_i = \beta_0 + \beta_1G_i + \epsilon_i$

- Model 2 : $y_i = \beta_0 + \beta_1H_i + \epsilon_i $

- Model 3 : $y_i = \beta_0 + \beta_1G_i + \beta_2H_i + \epsilon_i$

The likelihood corresponding to a model is given by : $l(\theta) = \sum$
Then the log-likelihood is : $\prod$

## C++ function 
```{Rcpp arma_logL}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::export]]
double arma_logL(arma::mat X, arma::vec y, arma::vec beta) {
  int n = y.n_elem;
  int p = X.n_cols;
  double logL = 0;
  for(int i = 0; i < n; i++) {
      double Xbeta_part = 0;
    for(int j = 0; j < p; j++) {
      Xbeta_part += X(i,j)*beta(j);
    }
    logL += R::dnorm(y[i], Xbeta_part, 1, 1);
  }
  return logL;
}
```

## R function 
```{r R_logL}
R_logL <- function(X,y,beta) {
  n <- length(y)
  p <- ncol(X)
  logL <- 0
  for (i in 1:n) {
    Xbeta_part <- 0
    for(j in 1:p) {
      Xbeta_part <- Xbeta_part + X[i,j]*beta[j]
    }
    logL <- logL + dnorm(x=y[i], mean=Xbeta_part, sd=1, log=T)
  }
  return(logL)
}
```

## Comparison of compilation times
```{r logL-benchmark}
# Trees data-set
y <- log(trees$Volume)
X <- cbind(1, log(trees$Girth), log(trees$Height))
# fit model y ~ X
arma_mod <- arma_fastLm(X,y)
R_mod <- R_fastLm(X,y)
# Call C++ function
Rcpp::sourceCpp("Code/arma_logL.cpp")
# Benchmark
Benchmark <- benchmark(
  "arma_logL" = {arma_logL(X,y,arma_mod$coefficients)},
  "R_logL" = {R_logL(X,y,R_mod$coefficients)},
  replications=100,
  columns = c("test", "elapsed", "relative"))
```

```{r logL-benchmark-results, echo=F}
library(dplyr)
knitr::kable(Benchmark,digits=3, booktabs=TRUE) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"),
		                          full_width=FALSE, position = "left")
```

## Representation of results 

```{r logL-results, echo=F}
# Trees data-set
y <- log(trees$Volume)
# fit models y ~ X
## model 1
X1 <- cbind(1, log(trees$Girth))
arma_mod1 <- arma_fastLm(X1,y)
R_mod1 <- R_fastLm(X1,y)
arma_logL1 <- arma_logL(X1,y,arma_mod$coefficients)
R_logL1 <- R_logL(X1,y,R_mod$coefficients)
## model 2
X2 <- cbind(1, log(trees$Height))
arma_mod2 <- arma_fastLm(X2,y)
R_mod2 <- R_fastLm(X2,y)
arma_logL2 <- arma_logL(X2,y,arma_mod$coefficients)
R_logL2 <- R_logL(X2,y,R_mod$coefficients)
## model 3
X3 <- cbind(1, log(trees$Girth), log(trees$Height))
arma_mod3 <- arma_fastLm(X3,y)
R_mod3 <- R_fastLm(X3,y)
arma_logL3 <- arma_logL(X3,y,arma_mod$coefficients)
R_logL3 <- R_logL(X3,y,R_mod$coefficients)
logL <- data.frame(arma_logL=c(arma_logL1,arma_logL2,arma_logL3), R_logL=c(R_logL1,R_logL2,R_logL3))
rownames(logL) <- c("model 1","model 2","model 3")
knitr::kable(logL,digits=3, booktabs=TRUE, row.names = T) %>%
		kableExtra::kable_styling(latex_options=c("HOLD_position","striped"),
		                          full_width=FALSE,position = "left")
```

